What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-09 android-stacks.tex
Consistency fixes

In this work, we perform the first large scale analysis of Java stack traces and
pinpoint how they can reveal bad programming practices. We mined and analyzed
the stack traces embedded in all issues of Java projects available GitHub.
Overall, our research set includes 356,057 issues define at 16,836 projects
(from which 28,800 stack traces were extracted). From this set, the stack traces
of 482 Android projects were investigated in more detail, in combination with
source code and bytecode analysis. In this study some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions) - revealing that Java hybrid exception
model is not fully used according to its purpose;  undocumented runtime
exceptions signaled by third party code - which makes almost impossible for
library clients to protect against such exceptions; and a high prevalence of
java.lang exceptions reported on issues - representing approximately 50\% of the
analyzed issues.

@@

%\todo{This section feels a bit disconnected.}
Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. 
% To guide our investigation, we compile the guidelines
A compiled list of the guidelines related to exception types and exception propagation in Java the following
\footnote{We also compiled guidelines related to exception handling but they are
out of the scope of this paper.}:

@@

\textbf{I-Use checked exceptions to represent recoverable
conditions}(~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The developer should use checked exceptions for conditions from which the caller
is expected to recover. By confronting the API user with a checked exception,
the API designer is telling the client to handle the exceptional condition. The
client can explicitly ignore the exception (swallowing, or converting it to
other type) at the expense of the program's robustness~\cite{gosling2000java}.

\textbf{II-Error represents an unrecoverable condition detected by the JVM which
should not he handled}~\cite{gosling2000java}. It results from failures detected
by the Java Virtual Machine which indicate resource deficiencies, invariant
failures or other conditions that make impossible the program to recover.

@@

\textbf{III-A method should throw exceptions that precisely define the
exceptional condition}~\cite{gosling2000java,bloch2008effective}. To do so,
developers should either try to reuse the exception types already defined in the
JVM or they should create a specific exception. Throwing general types such as a
pure Exception or a RuntimeException is hence considered a bad practice.

@@

%\noindent\emph{Exception Documentation}
\textbf{IV-Document all exceptions thrown by a
method}~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to programming
mistakes} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

These best practices guided our exploratory study described next
. Based on information extracted from stack traces (complemented by
bytecode and source-code analysis), we investigate whether stack characteristics
can reveal whether these practices have been obeyed. 
%Next section discusses the
%study procedure and the practices' misuses that emerged from this study.

@@

  \item[RQ3] What patterns of exception (mis)-use emerge from the stack trace
    analysis?
    %~\todo{Perhaps rephrase this to: Are the identified best practices followed
    %in practice?}

@@

We begun our investigation using the dataset provided by the GHTorrent
project~\cite{Gousi13}, an off-line mirror of the data offered through the
Github API. The Github API data come in two forms; a streaming data flow lists
events, such as reporting an issue, happening on repositories in real time,
while a static view contains the current state of entities. The data is stored
in unprocessed format, in a MongoDB database, while metadata is extracted and
stored in a relational database. The GHTorrent dataset covers a broad range of
development activities on Github, including issues. The project has been
collecting data since February 2012. Up to Dec 2013, when we queried it to
retrieve our bootstrapping sample which included: 356.057 issues from 16.836
Java projects (step 1).\footnote{A Java project is specified in GitHub with a
specific label.}

After obtaining the bootstrap issue sample from GHTorrent, each issue was
processed by ExceptionMiner tool developed in this work which extracted every
exception defined on issues or on issue comments (step 2). Since this initial
set of Java project is too diverse (it includes libraries, web-based or
stand-alone applications), in the second phase of this study we restrict the
analysis domain to Android projects (step 3). This decision was driven by the
fact that most open source Android applications are known~\cite{ref} for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.

@@

\noindent\emph{Kinds of Analyzed Issues.} Issues on Github are different from
issues on dedicated bug tracking tools such as Bugzilla and Jira. The most
important difference is that there are no predefined fields for bug reporters to
select from (e.g. severity and priority); Github uses instead a more open ended
tagging system. Repositories are offered a pre-defined set of labels, but
repository owners can modify them at will. An issue can have an arbitrary set of
labels attached. Hence, this study does not restrict the analysis of stack
traces defined on defect-issues, based on the assumption that: regardless the
type of issue where its is defined the stack trace information contain relevant
information concerning the exception use. Moreover, issue and pull requests are
dual on Github; all pull requests have a corresponding ``backing'' issue - which
are automatically generated. Therefore, we needed to exclude pull request based
issues from our analysis. Finally, the Github API allows the automated
generation of issues for specific repositories, which automated tools often use
to report crashes. In some cases, this led to high number of issues that
included stack traces. We identified and filtered those cases out (e.g.,
pullWifi project was responsible for almost 50\% of all issues created on
GitHub, it was removed from our analysis and also from GitHub).

@@

Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. selendroid, roboeletric -- tools to support the testing of Android apps).
The filtered set consisted of 482 apps, from which approximately 50\% are also
hosted in Google Market Place. For each one of the analyzed projects, we
downloaded the source code and using custom scripts, we extracted their package
hierarchy (recursive list of of Java packages) and their custom exception types
as described next.

@@

\noindent\emph{Exception Classifier} The next step in the ExceptionMiner process
is classifying exceptions as either checked exceptions, runtime exceptions or an
errors. The exception classifier is based on bytecode analysis (based on Design
Wizard~\cite{}) that walks up the type hierarchy of a Java exception until it
reaches a base exception type.

@@

Based on the assumption that regardless the issue type, every exception stack
trace contains relevant information concerning the exception structure of
projects analyzed we opted for not restricting the analysis only on defect
issues~\footnote{We conducted the same analysis on the defect issues and the top
exceptions are similar to the ones mentioned on defect issues. Due to space
limitation we limit to present the general analysis here. More detailed analysis
on the defect issues can be found here:
\url{www.dimap.ufrn.br/~roberta/icsm2014}}.

@@

For each exception we could discover the number of times it occurred on stacks,
the number of distinct projects such stacks were defined (and consequently its
popularity) and the of times it was signaled by different signaler types. We
could calculate the popularity of each root cause, considering the number of
distinct projects they were found and the whole set of projects analyzed with at
least one stack trace (482). Table~\ref{tab:toptenandroid} and
Figure~\ref{fig:androidsignaler} presents the mined data.

@@

We could observe in this study that the NullPointerException and other
other implicitly signaled exceptions represented most of the exceptions reported
on issues in both Java and Android subset. For such exceptions, which represent
programming bugs or resource limitations, there is usually no proper handling
besides presenting an error message to the user and restarting the application -
only high fault tolerant systems need to provide solutions to handle them.

@@

From Tables~\ref{tab:toptenjava} and ~\ref{tab:toptenandroid} we can observe that,
in both Java and Android related issues, direct instances
java.lang.RuntimeException were thrown  in 10,94\% of Java repositories and
18,88\% of Android repositories respectively. 
In the Android repositories, most of such exceptions  were  thrown by the Android 
platform/os (167 out of 232).

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

Runtime	&	1075	&	56	&	1446	&	374	&	10	&	2961	\\
Error	&	144	&	34	&	168	&	105	&	9	&	460	\\
Checked	&	110	&	230	&	139	&	145	&	12	&	636	\\
Throwable	&	0	&	0	&	1	&	0	&	0	&	1	\\
Undefined	&	1	&	0	&	6	&	8	&	0	&	15	\\
    \hline
All		& 1330	&	320	&	1760	&	632	&	31	&	4073	\\
    \hline
  \end{tabular}
\caption{Types of Root Exceptions.}
  \label{fig:typeroottab}
\end{table}

@@

On the other hand, we could observe that most of the exceptions signaled by the set of 
libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax - referred in
this work as libcore) were checked exceptions. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

@@

%\noindent\emph{3) Odd Wrappings}
\subsection{Odd Wrappings}
Table~\ref{wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions where indeed wrapped in runtime exceptions or even errors, while most
of them were not wrapped along the stack trace.

@@

 Checked    &   -  & 422  & 117 \\
     Error    &   -   & 381  & 141 \\
   Checked &  Runtime   & 109  & 70  \\
    Checked   & Checked & 98   & 42  \\
   Error   &  Runtime   & 55   & 38  \\
   Runtime &  Checked   & 22   & 12  \\
    Error     & Error   & 15   & 14  \\
  Undefined    &  -  &  15   & 10  \\
   Runtime & Error     &  13   & 6 \\
  Error   & Checked   &  8    & 7 \\
    Checked &Error     &  7    & 4 \\
    Runtime &Throwable &  3    & 1 \\
   Runtime & Undefined &  3    & 3 \\
   Throwable    & -  &  1    & 1 \\
   Error   & Undefined &  1    & 1 \\
    \hline
  \end{tabular}
\caption{Kinds of wrappings found on the stacks of 482 ANDROID projects found on Github}
\label{fig:wrappingandroid}
\end{table}

@@

We could also observe that in 205 stacks found in Android repository the exception was wrapped 
more then once. Figure ~\ref{sizewrapbhists}  illustrates the number of wrappings per stack, both
 in Java and Android studied repositories. One interesting multiple-wrapping found in Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

@@

\section*{Acknowledgment} This work is partially supported by: CNPq - Proc.
484209/2013-2 and the NWO TestRoots project (639.022.314).

>>>>

In this work, we perform the first large scale analysis of Java stack traces and
pinpoint how they can reveal bad programming practices. We mined and analyzed
the stack traces embedded in all issues of Java projects available GitHub.
Overall, our research set includes 356,057 issues define at 16,836 projects
(from which 28,800 stack traces were extracted). From this set, the stack traces
of 482 Android projects were investigated in more detail, in combination with
source code and bytecode analysis. In this study some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions), revealing that Java hybrid exception
model is not fully used according to its purpose;  undocumented runtime
exceptions signaled by third party code, which makes almost impossible for
library clients to protect against such exceptions; and a high prevalence of
java.lang exceptions reported on issues, representing approximately 50\% of the
analyzed issues.

@@

Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. 
A compiled list of the guidelines related to exception types and exception
propagation in Java the following: \footnote{We also compiled guidelines related
to exception handling but they are out of the scope of this paper.}

@@

\textbf{I-Use checked exceptions to represent recoverable
conditions} (\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective})
The developer should use checked exceptions for conditions from which the caller
is expected to recover. By confronting the API user with a checked exception,
the API designer is telling the client to handle the exceptional condition. The
client can explicitly ignore the exception (swallowing, or converting it to
other type) at the expense of the program's robustness~\cite{gosling2000java}.

\textbf{II-Error represents an unrecoverable condition detected by the JVM which
should not he handled} (\cite{gosling2000java}). It results from failures detected
by the Java Virtual Machine which indicate resource deficiencies, invariant
failures or other conditions that make impossible the program to recover.

@@

\textbf{III-A method should throw exceptions that precisely define the
exceptional condition} (\cite{gosling2000java,bloch2008effective}). To do so,
developers should either try to reuse the exception types already defined in the
JVM or they should create a specific exception. Throwing general types such as a
pure Exception or a RuntimeException is hence considered a bad practice.

@@

%\noindent\emph{Exception Documentation}
\textbf{IV-Document all exceptions thrown by a
method} (\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly signaled runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to programming
mistakes} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of the exceptions the method can throw. If the developer fails to
do follow this practice (specially when developing library code) it will be
difficult or even impossible for the caller to make effective use of such method
~\cite{wirfs2006toward, bloch2008effective}.

These best practices guided our exploratory study described next.
Based on information extracted from stack traces (complemented by
bytecode and source-code analysis), we investigate whether stack characteristics
can reveal whether these practices have been obeyed. 
%Next section discusses the
%study procedure and the practices' misuses that emerged from this study.

@@

  \item[RQ3] What patterns of exception (mis)-use emerge from the stack trace
    analysis?

@@

We begun our investigation using the dataset provided by the GHTorrent
project~\cite{Gousi13}, an off-line mirror of the data offered through the
Github API. The Github API data come in two forms; a streaming data flow lists
events, such as reporting an issue, happening on repositories in real time,
while a static view contains the current state of entities. The data is stored
in unprocessed format, in a MongoDB database, while metadata is extracted and
stored in a relational database. The GHTorrent dataset covers a broad range of
development activities on Github, including issues. The project has been
collecting data since February 2012. Up to Dec 2013, when we queried it to
retrieve our bootstrapping sample, it included 356.057 issues from 16.836
Java projects (step 1).\footnote{A Java project is specified in GitHub with a
specific label.}

After obtaining the bootstrap issue sample from GHTorrent, each issue was
processed by ExceptionMiner tool developed in this work which extracted every
exception defined on issues or on issue comments (step 2). Since this initial
set of Java project is too diverse (it includes libraries, web-based or
stand-alone applications), in the second phase of this study we restrict the
analysis domain to Android projects (step 3). This decision was driven by the
fact that most open source Android applications are known for dealing
with several sources of exceptions due to multi-threaded execution and
interaction with input/output resources; moreover they share a common
well-defined platform and a limited set of libraries.

@@

\noindent\emph{Kinds of Analyzed Issues.} Issues on Github are different from
issues on dedicated bug tracking tools such as Bugzilla and Jira. The most
important difference is that there are no predefined fields for bug reporters to
select from (e.g. severity and priority); Github uses instead a more open ended
tagging system. Repositories are offered a pre-defined set of labels, but
repository owners can modify them at will. An issue can have an arbitrary set of
labels attached. Hence, this study does not restrict the analysis of stack
traces defined on defect-issues, based on the assumption that: regardless the
type of issue where its is defined the stack trace information contain relevant
information concerning the exception use. Moreover, issue and pull requests are
dual on Github; all pull requests have a corresponding ``backing'' issue which
are automatically generated. Therefore, we needed to exclude pull request based
issues from our analysis. Finally, the Github API allows the automated
generation of issues for specific repositories, which automated tools often use
to report crashes. In some cases, this led to high number of issues that
included stack traces. We identified and filtered those cases out (e.g.,
the \textsf{pullwifi} project was responsible for almost 50\% of all issues created on
GitHub, it was removed from our analysis and also from GitHub).

@@

Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. \textsf{selendroid}, \textsf{roboeletric} --- tools to support the testing of Android apps).
The filtered set consisted of 482 apps, from which approximately 50\% are also
hosted in Google Market Place. For each one of the analyzed projects, we
downloaded the source code and using custom scripts, we extracted their package
hierarchy (recursive list of of Java packages) and their custom exception types
as described next.

@@

\noindent\emph{Exception Classifier} The next step in the ExceptionMiner process
is classifying exceptions as either checked exceptions, runtime exceptions or an
errors. The exception classifier is based on bytecode analysis (based on Design
Wizard~\cite{Brunet09}) that walks up the type hierarchy of a Java exception
until it reaches a base exception type.

@@

Based on the assumption that regardless the issue type, every exception stack
trace contains relevant information concerning the exception structure of
projects analyzed we opted for not restricting the analysis only on defect
issues.\footnote{We conducted the same analysis on the defect issues and the top
exceptions are similar to the ones mentioned on defect issues. Due to space
limitation we limit to present the general analysis here. More detailed analysis
on the defect issues can be found here:
\url{www.dimap.ufrn.br/~roberta/icsm2014}}

@@

For each exception we could discover the number of times it occurred on stacks,
the number of distinct projects such stacks were defined (and consequently its
popularity) and the of times it was signaled by different signaler types. We
could calculate the popularity of each root cause, considering the number of
distinct projects they were found and the whole set of projects analyzed with at
least one stack trace (482). Table~\ref{tab:toptenandroid} 
%and Figure~\ref{fig:androidsignaler}
presents the mined data.

@@

We could observe in this study that the NullPointerException and other
other implicitly signaled exceptions represented most of the exceptions reported
on issues in both Java and Android subset. For such exceptions, which represent
programming bugs or resource limitations, there is usually no proper handling
besides presenting an error message to the user and restarting the application ---
only high fault tolerant systems need to provide solutions to handle them.

@@

From Tables~\ref{tab:toptenjava} and~\ref{tab:toptenandroid} we can observe that,
in both Java and Android related issues, direct instances
java.lang.RuntimeException were thrown  in 10,94\% of Java repositories and
18,88\% of Android repositories respectively. 
In the Android repositories, most of such exceptions  were  thrown by the Android 
platform/OS (167 out of 232).

@@

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{tab:typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories. As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions. After filtering all the exceptions implicitly signaled by 
JVM (due to programming mistakes) and inspecting the signaler methods of such exceptions (i.e.,  direct an inderect 
instances of java.lang.RuntimeException). We could observe observe that only 1 method out of 118 inspected signalers
 (i.e., 0,8%) documented the explicitly thrown runtime exception in a Javadoc comment; and none of these methods
included the runtime exception  (reported on the stack) as part of the exception interface of the method (i.e., using 
throws clause on method signature). This result is aligned with the results of other  study conducted by from 
Sacramento et al ~\cite{sacramento2006unchecked} which observed that the
runtime exceptions in .Net programs are most often not documented.

@@

Runtime	&	1075	&	56	&	1446	&	374	&	10	&	2961	\\
Error	&	144	&	34	&	168	&	105	&	9	&	460	\\
Checked	&	110	&	230	&	139	&	145	&	12	&	636	\\
Throwable	&	0	&	0	&	1	&	0	&	0	&	1	\\
Undefined	&	1	&	0	&	6	&	8	&	0	&	15	\\
    \hline
All		& 1330	&	320	&	1760	&	632	&	31	&	4073	\\
    \hline
  \end{tabular}
\caption{Types of root exceptions.}
  \label{tab:typeroottab}
\end{table}

@@

On the other hand, we could observe that most of the exceptions signaled by the set of 
libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax -- referred in
this work as libcore) were checked exceptions. It is pointed as a good practice
for libraries (see Section~\ref{sec:best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

@@

%\noindent\emph{3) Odd Wrappings}
\subsection{Odd Wrappings}
Table~\ref{tab:wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions where indeed wrapped in runtime exceptions or even errors, while most
of them were not wrapped along the stack trace.

@@

 Checked    &   -  & 422  & 117 \\
     Error    &   -   & 381  & 141 \\
   Checked &  Runtime   & 109  & 70  \\
    Checked   & Checked & 98   & 42  \\
   Error   &  Runtime   & 55   & 38  \\
   Runtime &  Checked   & 22   & 12  \\
    Error     & Error   & 15   & 14  \\
  Undefined    &  -  &  15   & 10  \\
   Runtime & Error     &  13   & 6 \\
  Error   & Checked   &  8    & 7 \\
    Checked &Error     &  7    & 4 \\
    Runtime &Throwable &  3    & 1 \\
   Runtime & Undefined &  3    & 3 \\
   Throwable    & -  &  1    & 1 \\
   Error   & Undefined &  1    & 1 \\
    \hline
  \end{tabular}
\caption{Kinds of wrappings found on the stacks of 482 ANDROID projects found on Github}
\label{tab:wrappingandroid}
\end{table}

@@

We could also observe that in 205 stacks found in Android repository the exception was wrapped 
more then once. Figure ~\ref{fig:sizewrapbhists}  illustrates the number of wrappings per stack, both
 in Java and Android studied repositories. One interesting multiple-wrapping found in Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

@@

\section*{Acknowledgment} This work is partially supported by: CNPq -- Proc.
484209/2013-2 and the NWO TestRoots project (639.022.314).
